Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-08 paper/main.tex
Updates

As the automated algorithms can only recover the CPDAG from data, and our method
recovers the DAG. To compare them fairly, we computed the SHD and SID for all
possible orientations of the CPDAG and show them in the Figure.

To use our proposed approach, we need an expert to tell us the direction of the
edges. We simulated an expert using this assisted model construction approach,
we start with an empty model and take a greedy approach to add edges to this
model. At any point, we select the pair of variables that has significant
correlation and the highest unexplained correlation, i.e., highest conditional
association. We then use an oracle to decide the direction of the edge between
the variables. Given an accuracy of the oracle, $ \alpha $, the oracle either
gives the correct direction, incorrect direction, or returns None representing
that it does not know the direction.

@@

If the oracle returns None for any pair of variables, an edge between this pair
is not suggested to the oracle in future iterations. We also make sure that
only edges which do not form a cyclic in the graph are used to select the next
potential edge. After adding each edge, we also check if the p-value of any
existing edge shows independence. If that happens we remove that edge and
blacklist that edge, i.e., that edge is not prompted again during the run of
the algorithm. We repeat this procedure till the model explains all
correlations between the variables. There is a possibility that this greedy
expert gets stuck at incorrect graph structures as it does not do any
backtracking to fix its earlier mistakes. This procedure is outlined in the
Algorithm 1.
\todo[inline]{Show an algorithm of how this expert method is working}

\begin{verbatim}
	Input: Dataset $ D $ on variables $ \bm{X} $
	Output: Learned DAG G.

	function compute_effects(G):
		pvalues <- dict()
		effects <- dict()
		for X_i, X_j in X:
			if edge between X_i, X_j:
				p_values[(X_i, X_j)] <- pvalue(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
				effects[(X_i, X_j)] <- \phi(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
			else:
				p_values[(X_i, X_j)] <- pvalue(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
				effects[(X_i, X_j)] <- \phi(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
		return p_values, effects

	$ G $ <- Empty Graph
	p_values, effects <- compute_effects(G)
	while (p_values > p_thres, 
\end{verbatim}

\begin{algorithm}
	\KwData{Data set $ D $ on variables $ \bm{X} $, pthres, effectthres}
	\KwResult{The learned DAG: $ G $}
	\BlankLine
	$ G \leftarrow $ Empty Graph
\end{algorithm}

>>>>

\todo[inline]{Add equation for data generation method once the plot is finalized}

As the automated algorithms can only recover the CPDAG from data, and our
method recovers the DAG. To compare them fairly, we computed the SHD and SID
for all possible orientations of the CPDAG to show the best and worst case scenarios.

To use the \textit{ExpertInLoop} algorithm, we need two oracles: a d-separation
oracle and an ancestral oracle. We use a conditional independence test (partial
correlation if using linear data) for determining the d-separation with a
p-value threshold of $ 0.05 $. For the ancestral oracle, we simulated the
strong ancestral oracle. We vary the accuracy of our oracle to test the
performance for varying oracle accuracies. Given an accuracy level $ \alpha $
of the oracle, the oracle returns:

@@

If the oracle returns None for any pair of variables, an edge between this pair
is not suggested to the oracle in future iterations. We also make sure that
only edges which do not form a cyclic in the graph are used to select the next
potential edge. There is a possibility that this greedy expert gets stuck at
incorrect graph structures as it does not do any backtracking to fix its
earlier mistakes. This procedure is outlined in the Algorithm 1.
\todo[inline]{Show an algorithm of how this expert method is working}

% \begin{verbatim}
% 	Input: Dataset $ D $ on variables $ \bm{X} $
% 	Output: Learned DAG G.
% 	
% 	function compute_effects(G):
% 		pvalues <- dict()
% 		effects <- dict()
% 		for X_i, X_j in X:
% 			if edge between X_i, X_j:
% 				p_values[(X_i, X_j)] <- pvalue(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
% 				effects[(X_i, X_j)] <- \phi(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
% 			else:
% 				p_values[(X_i, X_j)] <- pvalue(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
% 				effects[(X_i, X_j)] <- \phi(X_i, X_j, pa_{\bar{G}}(X_i) \cup pa_{\bar{G}}(X_j))
% 		return p_values, effects
% 	
% 	$ G $ <- Empty Graph
% 	p_values, effects <- compute_effects(G)
% 	while (p_values > p_thres, 
% \end{verbatim}
% 
% \begin{algorithm}
% 	\KwData{Data set $ D $ on variables $ \bm{X} $, pthres, effectthres}
% 	\KwResult{The learned DAG: $ G $}
% 	\BlankLine
% 	$ G \leftarrow $ Empty Graph
% \end{algorithm}
